k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-02-01 paper.Rmd
minor

    When assigning stimuli to sets in experimental psychology, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) that simultaneously maximizes the diversity (corresponding to between-set similarity) and the dispersion (corresponding to pairwise stimulus dissimilarity within sets). We provide a first independent assessment of BILS and extend it in three ways. First, by prefixing an integer linear programming algorithm, our extension ensures that a globally optimal value of dispersion is always returned; without it, it is unknown if the BILS's result is optimal. Second, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, for example to prevent specific pairs of stimuli from appearing in the same condition. Third, our extension allows that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. Two simulation studies provide an extensive evaluation of the original BILS and our extension. Using the optimal approach as benchmark, we show that the original BILS identifies optimal dispersion values for much larger data sets than previously analyzed, but we also highlight conditions where it fails to do so reliably. A practical example shows how the BILS can be used for stimulus assignment in experimental psychology. The original BILS as well as our extension are both freely available via the open source R package anticlust.

@@

In within-subjects experiments, participants are presented with several stimulus sets that are affiliated with the different experimental conditions. Due to carry-over effects, showing the same stimuli in different conditions is oftentimes prohibited. This is evident in studies on memory, where studying a list of words in one conditions necessarily makes it impossible to repeat their study in a later phase; memory cannot be erased after an experimental block has been completed [@lahl2006]. While the experimental logic dictates that different lists must be used, the lists should be as parallel as possible with regard to variables that affect the response variable. @lintz2021 gave two reasons why it is important that study participants should be confronted with balanced lists: (a) if lists are not sufficiently parallel, the variance of within-subjects effect sizes due to the experimental manipulation will be inflated, reducing statistical power---if there is no proper statistical control of the stimulus materials, which is often the case [@judd2012]; (b) highly unbalanced sets may introduce secondary effects beyond the influence of the intrinsic stimulus features. That is, highly unequal lists may induce a strategic behavior shift between experimental blocks. Such a shift may for example result from suspicion, boredom or frustration, invalidating any observed differences between conditions. Importantly, this bias cannot be accounted for using statistical analysis.

>>>>

    When assigning stimuli to sets in experimental psychology, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) that simultaneously maximizes the diversity (corresponding to between-set similarity) and the dispersion (corresponding to pairwise within-set dissimilarity ). We provide a first independent assessment of BILS and extend it in three ways. First, by prefixing an integer linear programming algorithm, our extension ensures that a globally optimal value of dispersion is always returned; without it, it is unknown if the BILS's result is optimal. Second, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, for example to prevent specific pairs of stimuli from appearing in the same condition. Third, our extension allows that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. Two simulation studies provide an extensive evaluation of the original BILS and our extension. Using the optimal approach as benchmark, we show that the original BILS identifies optimal dispersion values for much larger data sets than previously analyzed, but we also highlight conditions where it fails to do so. A practical example shows how the extended BILS can be used for stimulus assignment in experimental psychology. The original BILS as well as our extension are both freely available via the open source R package anticlust.

@@

In within-subjects experiments, participants are presented with several stimulus sets that are affiliated with the different experimental conditions. Due to carry-over effects, showing the same stimuli in different conditions is oftentimes prohibited. This is evident in studies on memory, where studying a list of words in one conditions necessarily makes it impossible to repeat their study without being influenced by the earlier study phase; memory cannot be erased after an experimental block has been completed [@lahl2006]. While the experimental logic dictates that different lists must be used, the lists should be as parallel as possible with regard to variables that affect the response variable. @lintz2021 gave two reasons as to why it is important that study participants should be confronted with balanced lists: (a) if lists are not sufficiently parallel, the variance of within-subjects effect sizes due to the experimental manipulation will be inflated, reducing statistical power---if there is no proper statistical control of the stimulus materials, which is often the case [@judd2012]; (b) highly unbalanced sets may introduce secondary effects beyond the influence of the intrinsic stimulus features. For example, highly biased lists may induce a strategic behavior shift between experimental blocks, possibly due to suspicion, boredom or frustration. These effects invalidate any observed differences between conditions and importantly, this bias cannot be accounted for via statistical techniques.

Traditionally, assigning stimuli to experimental conditions was done manually [e.g., @schaper2019metamory] or randomly [e.g., @kroneisen2018], both of which methods do not guarantee that the stimulus sets will be similar [@papenberg2020; @cutler1981; @vancasteren2007; @armstrong2012]. Generally, automated approaches are preferred that quantify similarity on the basis of the stimulus features and use an algorithm to assign stimuli in such a way that this similarity is optimized [@brusco2019; @papenberg2020].
